Coalition Formation Algorithm of Prosumers in a Smart Grid Environment

### 2014-09-26 Article.tex
modif michel (in progress)

There seems to have a clear consensus on the benefit of having bidirectional communication flows between the prosumers and the grid, as demand side management relies on such an architecture \cite{WuHamedHuangBook2011}. Nevertheless as the number of active prosumers is expected to rise, it is safe to assume there is a need for more complex communication patterns (such as coalitions) that will help to decrease the communication burden and satisfy the multiple requirements of the power grid management. The formation of coalitions inside the smart grid environment could be applied to various types of agents such as self sufficient micro-grids \cite{Pahwa}, sizable and adjustable virtual power plants (VPP) \cite{Braun, Ramchurn}, or fleet of electric vehicles that back up the grid in some emergency situations (V2G) \cite{Ramchurn}. These are only a few use cases where the coalition of multiple agents can enhance the grid reliability and in the mean time reduce the communication burden.

@@

In section \ref{sec:model}, we consider that the agent's production depends on meteorological conditions (wind turbines and PV as generators), various energy mix and preferences, and their appliances (loads). We use meteorological data (see section \ref{subsec:ProsumerModel}) to account for seasonal and daily variations in the prosumer's energetic profiles as well as realistic geographical correlations between different agents. With these ingredients, we are able to run realistic simulations and record the different output profiles as time-series.

@@

On a narrower scale, the authors of \cite{Kota2011} study the formation of virtual power plants (VPP) composed of multiple self-interested DER. On the grid side, two requirements for the formation of virtual power plants are considered : the reliability of supply and the minimization of entities the grid has to deal with. From this, \cite{Kota2011} builds a pricing mechanism that encourages VPP to report true estimates of their aggregated production and penalizes prediction errors. A redistribution scheme of the VPP to the DER is also constructed such that the payoff allocation lies in the core of the game, meaning that no DER has an incentive to leave the coalition.

In this paper we ought to form coalitions utilities that depend on statistical properties of time-series values. The setting is thus similar with some financial studies on stock exchanges, where researchers tried to find relevant clusters of stocks based on their daily prices variations. The problem of clustering is usually approached by means of a similarity measure and completed by a clustering technique such as K-means or hierarchical clustering. Nevertheless, in \cite{Mantegna1999} the author introduced an approach where stocks time-series of their daily log returns, are organized in a graph such that stocks exhibiting similar price fluctuation patterns are close to each other. This closeness notion is formalized with a similarity measure based on Pearson correlation coefficient ($ d_{ij} = \dfrac{1}{2}\sqrt{2(1-\rho_{ij})} $ or $ d_{ij} = 1 - \rho_{ij}^{2} $ ) that enables to weight the edges of the graph.

@@

Presented in this way, the time-series clustering task seems very close to graph community detection. Communities in networks are indeed often seen as groups of nodes exhibiting high internal densities of links as well as a low density across communities \cite{Newmanb}. Although several techniques exists (based on different graph properties : modularity \cite{Newmanb}, edge betweenness \cite{Girvan2002}, spectrum of graph Laplacian \cite{Newman}), the clique percolation algorithm \cite{Lancichinetti} uses directly this observation and employs a greedy expansion rule based on a fitness function. When the algorithm stops, some coalitions may be partially overlapping, meaning that a node can belong to multiple communities. Detection of overlapping communities is actually a very active field of research, especially in social networks where a person might belong to several communities.

@@

Our major concern was to design a prosumer model with realistic patterns of consumption, production as well as realistic geographical correlations between them. We based our prosumer model on French's weather data from 2006 to 2012 sampled every 3 hours \cite{Infoclimat} (similar data are available for the United States covering 2010\cite{NCDC}).

As shown in the first part of figure 1, the first step consists in discretizing the studied zone around well chosen weather stations and gathering traces for these stations. We consider here that prosumers can only produce through wind turbines and photovoltaic panels (PV). On the other hand, we built their consumption patterns according to two major cycles :
\begin{itemize}
\item \textbf{Daily cycles} : Consumption is low during night, and higher during the day with two picks in the morning and evening. Some noise is added so that prosumers have similar but not identical cycles.
\item \textbf{Seasonal cycles} : Consumption is higher in the winter because of heating and low in the summer (air conditioning is not considered). Temperature traces were the principal ingredient for modeling these cycles.
\end{itemize}

@@

We thus collect for all chosen weather stations three kinds of data (average wind speed, cloudiness, and average temperature), which we will call a climate vector in the following. We consider that climate vectors are constant over their area, meaning that, if two agents are in the same area, they are exposed to the same climate vector.

@@

We thus used existing models enabling conversion between weather variables and output powers (see power curves in \cite{Kota2011} \cite{windturbinemodel}). \cite{Dans2007} provides also a convenient way of using cloudiness traces as realistic degradation factors in a \textit{"clear blue sky"} solar model, which enables us to rebuild realistic solar irradiance traces for stations that do not provide such information directly. Any agent is thus able to compute at any time how much he produces and consumes.

\subsection{Coalition formation model}\label{subsec:Coalition}
More formally, we denote by $ \mathcal{A} = \{ a_{1},...a_{N} \} $ the set of agents (the prosumers) and $ P_{i}(t) $ the instantaneous power value at agent i at instant t (its instantaneous production minus his instantaneous consumption, meaning that $ P_{i}(t) $ represents agent i available power at instant t). During the simulation (from $t_{0} $ to $ t_{K} $), all agents record their values with an hour time interval. As expected, the prosumer's time-series exhibit high seasonal patterns, but completely different from agent to agent because they depend both on the energetic mix and the habits. Nevertheless, these macro variations can hide the interesting information in the correlation coefficients and twist the rest of our process. Thereby, we remove these seasonal effects for all agents (see second block of figure \ref{Fig1}). We denote by $ \mathcal{T}_{i} = \{ P_{i}(t_{0}),...,P_{i}(t_{K}) \} $ the resulting time-series for  agent i, and by $ \mathcal{F}(P_{i}) $ the corresponding probability distribution.

We now extend these notations to any coalition $ s \subset \mathcal{A} $ : 
\begin{itemize}
\item $ P_{s}(t) = \sum_{i \in s} P_{i}(t) $
\item $ \mathcal{T}_{s} = \{ P_{s}(t_{0}),...,P_{s}(t_{K}) \} $ with $ \mathcal{F}(P_{s}) $ the corresponding probability distribution
\item A coalition S has the possibility to announce a contract value $ P_{s}^{CRCT} $ on the market.
\end{itemize}

>>>>

There seems to be a clear consensus on the benefit of having bidirectional communication flows between the prosumers and the grid, as demand side management relies on such an architecture \cite{WuHamedHuangBook2011}. Nevertheless as the number of active prosumers is expected to rise, it is safe to assume there is a need for more complex communication patterns (such as coalitions) that will help to decrease the communication burden and satisfy the multiple requirements of the power grid management. The formation of coalitions inside the smart grid environment could be applied to various types of agents such as self sufficient micro-grids \cite{Pahwa}, sizable and adjustable virtual power plants (VPP) \cite{Braun, Ramchurn}, or fleet of electric vehicles that back up the grid in some emergency situations (V2G) \cite{Ramchurn}. These are only a few use cases where the coalition of multiple agents can enhance the grid reliability and in the mean time reduce the communication burden.

@@

In section \ref{sec:model}, we consider that the agent's production depends on meteorological conditions (wind turbines and photovoltaic panels (PV) as generators), various energy mix and preferences, and their appliances (loads). We use meteorological data (see section \ref{subsec:ProsumerModel}) to account for seasonal and daily variations in the prosumer's energetic profiles as well as realistic geographical correlations between different agents. With these ingredients, we are able to run realistic simulations and record the different output profiles as time-series.

@@

On a narrower scale, \cite{Kota2011} study the formation, in a game theoretic setting, of virtual power plants (VPP) composed of multiple self-interested DER. Two requirements for the formation of virtual power plants are considered : the reliability of supply and the minimization of entities the grid has to deal with. From this, \cite{Kota2011} builds a pricing mechanism that encourages VPP to report true estimates of their aggregated production and penalizes prediction errors. A redistribution scheme of the VPP to the DER is also constructed such that the payoff allocation lies in the core of the game, meaning that no DER has an incentive to leave the coalition.

In this paper we do not consider stability against player defection but focus on a statistics-oriented definition of coalition stability. We ough indeed to form coalitions utilities that depend on statistical properties of time-series values. The setting is thus similar with some financial studies on stock exchanges, where researchers tried to find relevant correlated clusters of stocks based on their daily prices variations. There exists well-known algorithms such as K-means or hierarchical clustering that are traditionally used for such purposes. Nevertheless, despite their popularity, they do not appear particularly well suited for meaningfully translating complex correlation relationships between time series while clustering them, especially when these series exhibit complicated underlying patterns. In \cite{Mantegna1999} the author introduced an approach where stocks time-series of their daily log returns, are organized in a graph such that stocks exhibiting similar price fluctuation patterns are close to each other. This closeness notion is formalized with a similarity measure based on Pearson correlation coefficient ($ d_{ij} = \dfrac{1}{2}\sqrt{2(1-\rho_{ij})} $ or $ d_{ij} = 1 - \rho_{ij}^{2} $ ) that enables to weight the edges of the graph.

@@

Presented in this way, the time-series clustering task seems very close to graph community detection. Communities in networks are indeed often seen as groups of nodes exhibiting high internal densities of links as well as a low density across communities, and several topology oriented techniques for finding communities are present in the litterature (\cite{Newmanb} \cite{Girvan2002} \cite{Newman}). For our purpose where decorrelation is the closeness notion, such algorithms tend to have some difficulties because only a few inclusions of very high correlations can strongly affect the stability of a coalition. However, local cliques, where all nodes are linked, provide uncorrelated groups of nodes. There exists greedy heuristics based on cliques in the litterature such as clique percolation \cite{Lancichinetti} which realizes local optimizations of a fitness function and results in overlaping community structures. Detection of overlapping communities is actually a very active field of research, especially in social networks where a person might belong to several communities.

%Presented in this way, the time-series clustering task seems very close to graph community detection. Communities in networks are indeed often seen as groups of nodes exhibiting high internal densities of links as well as a low density across communities \cite{Newmanb}. Although several techniques exists (based on different graph properties : modularity \cite{Newmanb}, edge betweenness \cite{Girvan2002}, spectrum of graph Laplacian \cite{Newman}), the clique percolation algorithm \cite{Lancichinetti} uses directly this observation and employs a greedy expansion rule based on a fitness function. When the algorithm stops, some coalitions may be partially overlapping, meaning that a node can belong to multiple communities. Detection of overlapping communities is actually a very active field of research, especially in social networks where a person might belong to several communities.

@@

Our major concern was to characterize a prosumer behavior with realistic patterns of consumption, production as well as realistic geographical correlations between them. Only wind turbines and PV will be used as generators throughout this paper, and we denote by $ \mathcal{A} = \{ a_{1},...a_{N} \} $ the set of prosumers. To account for the production of a prosumer, we relied on French's weather data from 2006 to 2012 sampled every 3 hours \cite{Infoclimat} (similar data are available for the United States covering 2010\cite{NCDC}). As shown in the first part of figure 1, the first step consists in discretizing the studied zone around well chosen weather stations and gathering three kinds of traces (wind speed, cloudiness, and temperature) that we organize in a structure we will refer to as "climate vector" in the following. We consider that climate vectors are constant over their area, meaning that, if two agents are in the same area, they are exposed to the same climate vector. We used existing DER models enabling conversion between weather variables and output powers (see power curves in \cite{Kota2011} \cite{windturbinemodel}). \cite{Dans2007} provides also a convenient way of using cloudiness traces as realistic degradation factors in a \textit{"clear blue sky"} solar model, which enables us to rebuild realistic solar irradiance traces for stations that do not provide such information directly.

Consumption behaviors are characterized by a model we designed, which takes into account two major cycles :

\begin{itemize}
\item \textbf{Daily cycles} : Consumption is low during night, and higher during the day with two picks in the morning and evening. Some noise is added so that prosumers have similar but not identical cycles.
\item \textbf{Seasonal cycles} : Consumption is higher in the winter because of heating and low in the summer (air conditioning is not considered). Temperature traces are used for modeling these cycles.
\end{itemize}

@@

@@

Let $ P_{i}(t) $ be the instantaneous available power of agent i at instant t (his instantaneous production minus his instantaneous consumption). During the simulation (from $t_{0} $ to $ t_{K} $), all agents record their values with an hour time interval. As expected, the prosumer's time-series exhibit high seasonal patterns, but completely different from agent to agent because they depend both on the energetic mix and the habits. Nevertheless, these macro variations can hide the interesting information in the correlation coefficients and twist the rest of our process. Thereby, we remove these seasonal effects for all agents (see second block of figure \ref{Fig1}). We denote by $ \mathcal{T}_{i} = \{ P_{i}(t_{0}),...,P_{i}(t_{K}) \} $ the resulting time-series for  agent i, and by $ \mathcal{F}(P_{i}) $ the corresponding probability distribution.

\subsection{Coalition formation model}\label{subsec:Coalition}

We extend the agent notations to any coalition $ s \subset \mathcal{A} $ : 
\begin{itemize}
\item $ P_{s}(t) = \sum_{i \in s} P_{i}(t) $
\item $ \mathcal{T}_{s} = \{ P_{s}(t_{0}),...,P_{s}(t_{K}) \} $ with $ \mathcal{F}(P_{s}) $ the corresponding probability distribution
\item A coalition S has the possibility to announce a contract value $ P_{s}^{CRCT} $ on the market.
\end{itemize}
